---
title: "Coursera Machine Learning Project"
author: "Abdullah Albyati"
date: "August 31, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libraries}
library(caret)
library(dplyr)
library(randomForest)
library(rpart)
library(rpart.plot)
```

##Reading Data

```{r data, echo=TRUE}
set.seed(1234)
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "training.csv")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "testing.csv")
#data for partitioning
training <- read.csv("training.csv", na.strings=c("NA","","#DIV/0!"))
#testing data to validate 
validate <- read.csv("testing.csv")
```

##Cleaning and Preparing the Data
First we start with identifying near zero variance variables(NZV) and creating clean data sets without the NZV.
Then we remove the unclean sets.
```{r Cleaning-NVZ, echo=TRUE}
NZV <- nearZeroVar(training, saveMetrics = TRUE) 
trainingclean <- training[, !NZV$nzv]
validateclean <- validate[, !NZV$nzv]
rm(training)
rm(validate)
rm(NZV)
```

Our data sets has some columns that are unnecessary to our analysis we will remove these columns all together
```{r Cleaning-columns, echo=TRUE }
usless <- grepl("^X|timestamp|user_name", names(trainingclean))
training <- trainingclean[, !usless]
testing <- validateclean[, !usless]
rm(usless)
rm(trainingclean)
rm(validateclean)
dim(training)
dim(testing)
```

Last step in the cleaning process is removing all the columns that are made entirely of NA's

```{r Cleaning-NAColumns, echo=TRUE}
notneeded <- (colSums(is.na(training)) == 0)
training <- training[, notneeded]
testing <- testing[, notneeded]
rm(notneeded)
```

##Partitioning Training Set
In this section I will split the training data to 70% for training and 30% for validation  

```{r Spliting, echo=TRUE}
inTrain <- createDataPartition(training$classe, p = 0.70, list = FALSE)
validation <- training[-inTrain, ]
training <- training[inTrain, ]
rm(inTrain)
```

Now we have three clean data sets in the project environment with the following dimensions

```{r dimensions, echo=TRUE}
dim(testing)
dim(training)
dim(validation)
```

##Modeling the Data

Using Decision Tree model

```{r decision-tree, echo=TRUE}
tree <- rpart(classe ~ ., data = training, method = "class")
prp(tree)
```

Let's check the model against our validation data set to check for performance to include confusion matrix and accuracy

```{r validate-tree, echo=TRUE}
predictTree <- predict(tree, validation, type = "class")
confusionMatrix(validation$classe, predictTree)
accuracy <- postResample(predictTree, validation$classe)
OutOfSample <- 1 - as.numeric(confusionMatrix(validation$classe, predictTree)$overall[1])

```


